On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-22 paper/paper3d.tex
new eigenvector fig
starting another revision

\begin{abstract}
%Low mass, self-gravitating accretion disks can remain in a  
%quasi-steady, `gravito-turbulent' state provided that thermal 
%time-scales are sufficiently long, or equivalently for sufficiently
%weak turbulent stresses. Otherwise, numerical simulations show that
%disk fragmentation ensues. 
%{\bf
Low mass, self-gravitating accretion disks admit a quasi-steady,
`gravito-turbulent' state in which cooling balances turbulent
heating. However, numerical simulations show that gravito-turbulence
cannot be sustained when the cooling rate or  
turbulent viscosity is too large, resulting in disk fragmentation.     
%} 
We quantify the interpretation of this result as the
compromise of gravitational stability by: 
\begin{inparaenum}[1)] 
\item 
cooling, which reduces pressure support; and/or
\item 
viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence modeled as an effective
viscosity and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
%We suggest that realistic disk fragmentation can be
%physically attributed,  
%at least partially, to the removal of rotational stabilization against
%self-gravity by turbulence.   
We suggest turbulent stresses play an active role in realistic disk
fragmentation by removing rotational stabilization against 
self-gravity. 
%}
\end{abstract}
%we suggest turbulent stresses play an active, dynamical role in disk fragmentation by 
%removing rotational stabilization against self-gravity. 
%in the sense of small-scale transport

\section{Introduction}\label{intro} 
The gravitational instability (GI) of rotating disks    
plays a fundamental role in many astrophysical problems. In the context of 
gaseous protoplanetary disks (PPDs), GI has two applications. It can  
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion inwards \citep{armitage10,turner14}.   
GI may also lead to disk fragmentation, which has been invoked to 
explain giant planet formation at large radii \citep{helled14}. Any 
realistic study of these phenomena require direct   
numerical simulations. Nevertheless, physical insight can be  
obtained through analytical understanding of GI.  %physical insight

The standard metric for the (inverse) strength of disk self-gravity is the 
\emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). When $Q\leq 1$ the disk is formally unstable to local,
axisymmetric disturbances in a two-dimensional (2D, razor-thin),
inviscid and adiabatic (including isothermal) disk.

@@

\subsection{Cooling removes pressure support}
A net thermal loss\footnote{Isothermal disks implicitly assume 
  heating and cooling processes are exactly balanced at all
  times, so these do not have cooling in the current context.} (hereafter `cooling')
reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some pre-defined value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

@@

\subsection{Viscosity removes rotational support}
%there is also viscous damping which stabilizes
Viscous disks can develop GI even when $Q>1$. This is because
viscosity compromises rotational support against self-gravity for
long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect  
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
is a mechanism to enhance particle clumping for planetesimational formation 
%frag of dust 
\citep{youdin11}. Then it is not unreasonable to expect an analogous
fragmentation of gaseous disks due to viscosity.

>>>>

\begin{abstract}
Low mass, self-gravitating accretion disks admit quasi-steady,
`gravito-turbulent' states in which cooling balances turbulent
heating. However, numerical simulations show that gravito-turbulence
cannot be sustained beyond dynamical timescales when the cooling rate
or turbulent viscosity is too large, resulting in disk fragmentation.        
%We quantify the interpretation of this result as the
%compromise of gravitational stability by: 
We quantify the interpretation of this result as a 
secondary gravitational instability due to: 
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to
gravito-turbulence modeled as an effective 
viscosity and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial
%this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses play an active role in realistic disk
fragmentation by removing rotational stabilization against 
self-gravity. 
\end{abstract}

\section{Introduction}\label{intro} 
The gravitational instability (GI) of rotating disks    
plays a fundamental role in many astrophysical problems. In the context of 
gaseous protoplanetary disks (PPDs), GI has two applications. It can  
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion \citep{armitage10,turner14}.   
GI may also lead to disk fragmentation, which has been invoked to 
explain giant planet formation at large radii
\citep{helled14}. Studying these non-linear phenomena require direct      
numerical simulations. Nevertheless, physical insight can be  
obtained through analytical understanding of GI.

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), 
inviscid and adiabatic (including isothermal) disk, 
\begin{align} 
  s^2 = \kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|, 
\end{align}
relating the growth rate $s$ and the radial wavenumber $k$. When
$Q\leq 1$ the disk is formally unstable.

@@

\subsection{Cooling-driven gravitational instability}
A net thermal loss\footnote{Isothermal disks implicitly assume 
  heating and cooling processes are exactly balanced at all
  times, so these do not have cooling in the current context.} (hereafter `cooling')
reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some pre-defined value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

@@

\subsection{Viscosity-driven gravitational instability}
%there is also viscous damping which stabilizes
Viscous disks can develop GI even when $Q>1$. This is because
viscosity compromises rotational support against self-gravity for
long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect  
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
is a mechanism to enhance particle clumping for planetesimational formation 
%frag of dust 
\citep{youdin11}. Then it is not unreasonable to expect an analogous
fragmentation of gaseous disks due to viscosity.
